---
title: "Working with manual files to study disease networks with Omics Visualizer and the String App"
output: amyloidoses_networks
---
#Set working directory in R markdown
```{r setup}
knitr::opts_knit$set(root.dir = getwd())
```

#Initialize
Make sure the required packages are installed. If any package is missing you can always install it with the install.packages("packageName") command and load it with library(packageName).
```{r}
if(!"RCy3" %in% installed.packages()){
    install.packages("BiocManager")
    BiocManager::install("RCy3")
}
library(RCy3)

if(!"RColorBrewer" %in% installed.packages()){
    install.packages("RColorBrewer")
}
library(RColorBrewer)
help(package=RCy3)
```

##Check connection and close previous session if necessary
Make sure you have the latest version of Cytoscape installed and you have already started Cytoscape. If you want to close previous session uncomment the second line in the next block.
```{r}
cytoscapePing()
#closeSession(TRUE)
```

##Make sure you have the required Apps installed inside Cytoscape
For this to work you will need an Internet connection, as will be the case for the rest of this exercise
```{r}
installedCysApps <- getAvailableApps()
#remove prefix and suffix
installedCysApps <- sub("(name: )(.*)(, version:.*)", "\\2", installedCysApps, perl=TRUE)

if(!"stringApp" %in% installedCysApps) {
  installApp("stringApp")
}
if(!"enhancedGraphics" %in% installedCysApps) {
  installApp("enhancedGraphics")
}
if(!"Omics Visualizer" %in% installedCysApps) {
  installApp("Omics Visualizer")
}
```

#Load and Preprocess data
##Load network from table
In this exercise, we will load a data table with disease-protein association data, create a protein-disease network, retrieve STRING networks for the proteins associated with each disease seperately, and visualize disease-specific information onto the protein network using Omics Visualizer.

The data have been retrieved from AmyCo ([Nastou, et. al, 2019] (https://www.ncbi.nlm.nih.gov/pubmed/31094220)) a database of amyloidoses and diseases related to the amyloid deposition.

Please follow the Omics Visualizer and String App exercises before moving on.

```{r}
network_file <- paste(getwd(), "amyco_disease_protein_interactions.tsv", sep = '/')
Sys.sleep(0.5)
if(!file.exists(network_file)) {
  download.file("https://jensenlab.org/assets/amyco_disease_protein_interactions.tsv", network_file)
  
}
#columnTypeList: List of column types ordered by column index (e.g. "source,target,interaction,source attribute,target attribute,edge attribute,skip" or just "s,t,i,sa,ta,ea,x"):
network_load=paste0('network import file file="',network_file,'" columnTypeList="sa,s,t,ta,ea,ea,ea" firstRowAsColumnNames=true startLoadRow=1  delimiters="\\t"')
#remember to change the \\ to \ in cytoscape command line
Sys.sleep(1)


commandsRun(network_load)
#Did you change the working directory???
Sys.sleep(1)
suid_main <- getNetworkSuid()
```

##Clone network and Remove APP and PRNP
At this stage we want to remove two proteins that might introduce noise in our analysis. APP is associated with Alzheimer's disease and PRNP is associated with Creutzfeld-Jakob disease. So, we expect that the majority of high-scoring functional associations retrieved from STRING for these proteins will be associated with these highly prevalent diseases, and not the ones under study.
```{r}
setCurrentNetwork(suid_main)
#clone network
cloneNetwork(network=suid_main)
Sys.sleep(1)
#get suid
suid_no_app_prnp <- getNetworkSuid()
#select this network
setCurrentNetwork(suid_no_app_prnp)
Sys.sleep(1)
renameNetwork("All_interactions_no_app_prnp", network=suid_no_app_prnp)

#remove app and prp
selectNodes("APP,PRNP", by.col="name")
deleteSelectedNodes()
```

##Delete unnecessary network
```{r}
net_to_delete <- sub("(/.*/)(.*)", "\\2", network_file, perl=TRUE)
deleteNetwork(network=net_to_delete)
```

##Import attribute table with protein/disease labels
Here we load an attribute file, that will add a label on every node in our network. We will use those labels to select nodes associated with specific diseases later.
```{r}
attribute_file <- paste(getwd(), "protein_or_disease_amyco.txt", sep = '/')
if(!file.exists(attribute_file)) {
  download.file("https://jensenlab.org/assets/protein_or_disease_amyco.txt", attribute_file)
  
}

data <- read.delim(attribute_file, stringsAsFactors = FALSE)
#the key column from the attribute file. that will be used to map the data to the preloaded network
loadTableData(data,data.key.column="name")
```
##Load Omics Visualizer table
The same table that we used previously to load the network will also be used to visualize data with Omics Visualizer
```{r}
ov.cmd3 <- paste0('ov load file="',network_file,'" newTableName="All_data" firstRowAsColumnNames=true startLoadRow=1')
commandsRun(ov.cmd3)
```

#Create STRING networks
##Create subnetworks per disease
At this stage we will create one protein subnetwork for each disease and then query STRING to retrieve interactions between the proteins associated with each disease, as well as, the top 10 interactors each time. Then we will create special visualizations for each network using Omics Visualizer.

##Function for STRING queries
Before we move forward we will write a function to perform String Protein Queries.
This function can be used to perform any protein query when working with stringApp v1.5 and is just implemented for convienience. You can always write string queries the same way we did for omics visualizer above.
```{r}
# ------------------------------------------------------------------------------
#' @title String Protein Query
#'
#' @description Retrieves a string Network for a set of Proteins
#' @details Enter protein names or identifiers to query the STRING database for protein-protein interactions.
#' @param cutoff (optional) <BoundedDouble (0.0<=0.4<=1.0)>: The confidence score reflects the cumulated evidence that this interaction exists. Only interactions with scores greater than this cutoff will be returned
#' @param includesViruses (optional) true|false: By default, a query will search for identifiers in both the protein and virus databases. By changing this to 'false', only the protein database will besearched
#' @param limit (optional) <BoundedInteger (0<=10<=10000)>: The maximum number of proteins to return in addition to the query set
#' @param network (optional) network=current Specifies a network by name, or by SUID 
#' @param newNetName (required) newNetName=<String>: Name for the network to be created
#' @param query (required) <String>: Comma separated list of protein names or identifiers
#' @param species (optional) <String>: Species name. This should be the actual taxonomic name (e.g. homo sapiens, not human)
#' @param taxonID (optional) <int>: The species taxonomy ID. See the NCBI taxonomy home page for IDs. If both species and taxonID are set to a different species, the taxonID has priority.
#' @return None
#' @examples \dontrun{
#' 
#' }
#' @export
stringProteinQuery <-  
  function(cutoff=0.4,
  includesViruses="true",
  limit=0,
  network=NULL,
  newNetName=NULL,
  query=NULL,
  species="Homo sapiens",
  taxonID=9606) {
  if(is.null(query)){
    write("Please provide protein identifiers for the query",stderr())
  } else {
    cmd <- 
      paste0('string protein query cutoff=',
             cutoff,
             ' includesViruses=',
             includesViruses,
             ' limit=',
             limit,
             " network=",
             network,
             ' newNetName="',
             newNetName,
             '" query="',
             query,
             '" species="',
             species,
             '" taxonID=',
             taxonID
             )
    res <- commandsPOST(cmd)
    invisible(res)
  }
  }
```


The first step is to Select nodes with attribute "disease" and put them in a list
```{r}
diseases <- selectNodes("disease", by.col="type")
diseases_node_suids <- diseases$nodes
```

#Create disease subnetworks
Then for each disease create subnetworks
```{r}
network_names <- c()
string_network_names <- c()
#set flag for visual style
flag = 0
for (i in diseases_node_suids){
  #if you want to change preferred layout, do it through the gui
  layoutNetwork()
  #select disease nodes one by one
  selectNodes(i, preserve.current.selection=FALSE)
  #get disease name attribute to name the new network for each disease
  disease_name <- getTableValue('node',i, 'DISEASE_NAME')
  
  #Select first neighbors of disease node: these will be the proteins associated with each disease
  first_neighbors <- selectFirstNeighbors()
  #check if the disease is connected to any proteins (is not a singleton) after removing any proteins during preprocessing 
  if (length(first_neighbors$nodes) !=1){
    #Create network from selection, i.e. disease + proteins associated with the disease
    createSubnetwork(subnetwork.name = disease_name)
    Sys.sleep(0.5)
    #Delete the disease node from the network, to keep only protein nodes for the STRING protein query later
    selectNodes(i, preserve.current.selection=FALSE)
    deleteSelectedNodes()
    #get node names to make the query string for stringapp
    nodes <- getAllNodes()
    #create a comma separated string variable to store the protein names for the query
    query_string <- paste(nodes, collapse=",") 
    Sys.sleep(0.5)
    #run string protein query using the function to retrieve the top 10 interactors
    #alternatively uncomment the 2 lines of code below this command and run using that -- But don't do both!!!
    stringProteinQuery(includesViruses="false", limit=10, newNetName = disease_name, query=query_string)
    Sys.sleep(1)
    #string.cmd <- paste('string protein query cutoff=0.4 includesViruses=false limit=10 taxonID=9606 newNetName=\"',disease_name,'\" query=\"',query_string, '\"',sep='')
    #commandsRun(string.cmd)
    
    
    ##### VISUAL STYLE CREATION #####
    if(flag==0){
        # Create a style to apply to all networks
        #Sometimes the app creates one style for each network and not a generally applicable style (e.g. STRING style v1.5), so you have to make sure one visual style exists that can be applied to all networks
        copyVisualStyle(from.style=paste("STRING style v1.5 - ", disease_name, sep=''), to.style="AmyCo style")
        flag=flag+1
        Sys.sleep(0.5)
        #Remove string glass balls and fill colors
        deleteStyleMapping(style.name="AmyCo style", visual.prop="NODE_CUSTOMGRAPHICS_1")
        Sys.sleep(0.5)
        deleteStyleMapping(style.name="AmyCo style", visual.prop="NODE_FILL_COLOR")
        Sys.sleep(0.5)
        #set default node color to gray
        setNodeColorDefault('#C0C0C0', style.name ="AmyCo style")
        Sys.sleep(0.5)
        lockNodeDimensions(TRUE, style.name = "AmyCo style")
        Sys.sleep(0.5)
        #Set node size to 30
        setNodeSizeDefault(new.size=30, style.name = "AmyCo style")
        Sys.sleep(0.5)
    }
    ##### END OF VISUAL STYLE CREATION #####
    
    
    #load attribute file for each STRING network
    data <- read.delim(attribute_file, stringsAsFactors = FALSE)
    loadTableData(data,data.key.column="name",table.key.column="display name")
    #Apply the visual style you created above
    setVisualStyle("AmyCo style")
    Sys.sleep(1)
    #make nodes of query bigger
    for (i in nodes){
      setNodeSizeBypass(i, 50)
    }
    #connect OV table to network
    ov.cmd2 <-'ov connect mappingColNet="display name" mappingColTable="GENENAME"'
    commandsRun(ov.cmd2)
    Sys.sleep(1)
    #create custom color mapping and apply inner graphics
    #you can always use one of the palletes provided by the app.
    #e.g. ov.cmd4 <- 'ov viz apply inner discrete paletteName="viridis" paletteProviderName="Viridis"'
    #diseases are colored based on some common attributes in order to reduce the colors to 12
    ov.cmd4 <- 'ov viz apply inner discrete attributes="DISEASE_NAME" colorMapping="Immunoglobulin Light-chain Amyloidosis:#80b1d3,Localized insulin-derived Amyloidosis:#bebada,Apolipoprotein A-I associated Amyloidosis:#fb8072,Hereditary fibrinogen A alpha-chain Amyloidosis:#e31a1c,Hereditary lysozyme Amyloidosis:#ffffb3,Apolipoprotein A-II associated Amyloidosis:#fb8072,AH Amyloidosis:#80b1d3,AHL Amyloidosis:#80b1d3,Apolipoprotein C-II associated Amyloidosis:#fdb462,Apolipoprotein C-III associated Amyloidosis:#fdb462,Leukocyte chemotactic factor 2 Amyloidosis:#33a02c,Apolipoprotein A-IV associated Amyloidosis:#fb8072,AA Amyloidosis:#1F78B4,Wild-type transthyretin-related Amyloidosis:#b3de69,Wild type beta-2-microglobulin-related Amyloidosis:#fccde5,Calcifying Epithelial Odontogenic Tumor:#ffffb3,Variably Protease-Sensitive Prionopathy:#d9d9d9,Isolated Atrial Amyloidosis:#b3de69,Medin Amyloidosis:#b3de69,Enfuvirtide-induced Amyloidosis:#bebada,PrP Systemic Amyloidosis:#d9d9d9,Senile Seminal Vesicle Amyloidosis:#ffffb3,Hereditary beta-2-microglobulin-related Amyloidosis:#fccde5,Finnish type amyloidosis:#ccebc5,Transthyretin amyloidosis:#8dd3c7,Primary cutaneous amyloidosis:#bebada,Hereditary Cerebral Amyloid Angiopathy Icelandic Type:#bc80bd,APP-related cerebral amyloid angiopathy:#bc80bd,ITM2B-related cerebral amyloid angiopathy 1:#bc80bd,ITM2B-related cerebral amyloid angiopathy 2:#bc80bd"'
    commandsPOST(ov.cmd4)
    Sys.sleep(1)
    #the relationships between the proteins and the diseases are mapped on the outer part of the node. Label 3 (black color) indicates an amyloidogenic protein for the specified disease, Label 2 (gray color) indicates a co-deposited protein for the speficied disease, which also is an amyloidogenic protein in another disease and Label 1 (white color) indicates a co-deposited protein with no pre-recorded amyloidogenic properties in any other disease
    ov.cmd5 <- 'ov viz apply outer continuous colorMin="#d3d3d3" colorMax="#000000" colorMid="#a9a9a9" attributes="LABEL" rangeMax=3 rangeMid=2 rangeMin=1'
    commandsPOST(ov.cmd5)
    Sys.sleep(1)
    ######################################################################
    ########       This node segment is for edge filtering      ##########
    ########  Apply only if necessary, otherwise comment out    ########## 
    ######################################################################
    #apply edge filter to use experimental evidence only
    createColumnFilter(filter.name="experiments_only", column="stringdb::experiments", criterion=c(0,1), predicate="BETWEEN", type = "edges")
    #invert edge selection to select edges with no experimental evidence
    invertEdgeSelection()
    #delete selected edges
    deleteSelectedEdges()
    Sys.sleep(1)
    ######################################################################
    ###########              End of code segment               ###########
    ######################################################################
    string_network_names <- c(string_network_names, getNetworkName())
    #get network names in a list
    network_names <- c(network_names,disease_name)
    Sys.sleep(1)
  }
  ##Select main network
  setCurrentNetwork(network=suid_no_app_prnp)
}
```
##Save session
```{r}
saveSession(paste(getwd(),"one_network_per_disease.cys", sep = '/'))
```

##Apply Layout
These steps can be taken as a first effort to layout the network, but generally if the layout is applied manually the visualization is better.
Cose seems to work best amongst the automated layouts tested.
```{r}
network.list <- getNetworkList()
for (i in network.list){
  if (grepl("String Network",i)){
    setCurrentNetwork(i)
    #apply organic layout - set as default through the GUI
    layoutNetwork()
    #apply CoSE layout
    layoutNetwork("cose")
  }
}
```


```{r}
saveSession(paste(getwd(),"one_network_per_disease.cys", sep = '/'))
```